---
title: "Megan Lott Coding Exercise"
output:
  html_document:
    df_print: paged
---

We're going to learn to code by playing around with some of the data in the dslabs package. 

```{r}
library(dslabs)

#use the help function tosee what the dataset gapminder contains

help(gapminder)
#?gapminder would also work

#inspect the data 

str(gapminder)

summary(gapminder)

class(gapminder)

names(gapminder)

```

By inspecting the data, we can see that gapminder is a data frame that consists of demographic information of the world's counties. The information includes: infant mortality, life expectancy, fertility, populayion, gdp, continent, region. 

We want to look more closely at countries from Africa. We are going to extract that data using the subset() function in base R. 

```{r}

#assign only the African countries to new objects/variables 

africadata = subset(gapminder, continent == "Africa")
summary(africadata)


```

Now, we have only 2907 observations. We are interested in examining the infant mortality, life expectancy, and population of the countries in Africa. 

```{r}

#make two new variables: one that contains only infant_mortality and life_expectancy and one that contains only population and life_expectancy. The c() function might be useful to efficiently pull out the variables you want. 

africa_data_set1 = subset(africadata, select=c(infant_mortality, life_expectancy))
africa_data_set2 = subset(africadata, select=c(population, life_expectancy))

#You should have two new objects/variables with 2907 rows and two columns.
#NOTE: We no longer have the country information. 
#what are the units on infant mortality?

#Use the str, and summary commands to take a look at both variables.

str(africa_data_set1)
summary(africa_data_set1)


```


We are going to examine the data on infant mortality, life expectancy, and population by plotting this data. 

```{r}
#plot life expectancy as a function of infant mortality 

plot(africa_data_set1$infant_mortality, africa_data_set1$life_expectancy, xlab = "Infant Mortality Rate (deaths per 1,000 births)", ylab = "Life Expectancy (Years)", main = "Life Expectancy as a Function of Infant Mortality in African Countries")

#plot life expectancy as a function of population size 

plot(africa_data_set2$population, africa_data_set2$life_expectancy, log = "x", xlab = "Population (log)", ylab = "Life Expectancy (Years)", main = "Life Expectancy as a Function of Population in African Countries")
```

We see a negative correlation between infant mortality and life expectancy. We see a positive correlation between population size and life expectancy, but this data has streaks. **Why is this?**

We have different years for individual countries. Overtime, these countries increase in population size and in life expectancy. To see the relationship between the two variables in focus, we will tease out the data from a single year of interest. We will look at the year for which we have the most amount of data. 

```{r}

#Write some base R code that figures out which years have missing data for infant mortality. The is.na() function might be helpful. You can use the print() function to print the missing years to the console.


#check_na = is.na(africadata$infant_mortality[8])

#print(africadata$year[8])


years_missing_data = data.frame()

x = 2907

for(x in 1:2907){
if(is.na(africadata$infant_mortality[x] == "TRUE")){years_missing_data = rbind(years_missing_data, africadata$year[x])}}


#You should find that there is missing up to 1981 and then again for 2016. So we’ll avoid those years and go with 2000 instead. 

#create a new object by extracting only the data for the year 2000 from the africadata object. You should end up with 51 observations and 9 variables. Check it with str and summary

year_2000 = subset(africadata, year == 2000)

str(year_2000)
summary(year_2000)

```

Now, we can examine the relationship between infant mortality, life expectancy, and population in the Year 2000. 

```{r}

#use base R plotting again and do the same two plots again, this time only for the year 2000

plot(year_2000$infant_mortality, year_2000$life_expectancy, xlab = "Infant Mortality Rate (deaths per 1,000 births)", ylab = "Life Expectancy (Years)", main = "Life Expectancy vs. Infant Mortality in African Countries in the Year 2000")

plot(year_2000$population, year_2000$life_expectancy, log = "x", xlab = "Population (log)", ylab = "Life Expectancy (Years)", main = "Life Expectancy vs. Population in African Countries in the Year 2000")


```

We see that there is a negative correlation between infant mortality and life expectancy, but no noticeable correlation between population size and life expectancy. 

```{r}
# Use the lm function and fit life expectancy as the outcome, and infant mortality as the predictor. Then use the population size as the predictor. 

fit1 = lm(year_2000$life_expectancy ~ year_2000$infant_mortality)

summary(fit1)

fit2 = lm(year_2000$life_expectancy ~ year_2000$population)

summary(fit2)

```

The p-value for fit1 is 2.826e-08. There is a significant correlation between life expectancy and infant mortality. The p-value for fit2 is 0.6159. There is no significant correlation between life expectancy and population. 

